 

### 2021-07-14 vertical-thermal-review.Rmd
added some citations

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen $1.2^\circ$C since the 1850-1900 period [@rohdeBerkeleyEarthLand2020] and are expected to rise over pre-industrial levels of $1.6 - 4.3^{\circ}$C by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilitiesa; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilitiesa; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilitiesa]. Collectively, these impacts may lead to reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020]. The resulting feedbacks to the climate system, decreased C storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016], yet this influence remains extremely uncertain [REFS].

@@

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
Foliage distribution is structured in correspondence with the distribution of tree sizes within a forest, where leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs. `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`). 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In open canopy forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations followed by disturbance all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
In this review, we focus on growing season conditions (peak LAD) unless otherwise noted.

@@

**Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al. 2015 ; Slot et al. 2019; Chen et al. 2020).**
This is primarily driven by the greater light available to sun-leaves, that also have traits [Table 1; Table 2] allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions. Photosynthesis increases with irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$. Light saturation levels differ between leaves and whole canopies with leaves saturating at lower light level than canopies; while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy.

@@

In 1994, Knapp et al. showed that $T_{50}$ measured on chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions. In both species $T_{50}$ also tracked increased from June to August. Within species, @slotPhotosyntheticHeatTolerance2019a found that $T_{50}$ was slightly lower for shade than sun leaves for two of three tree species measured in Panama.

Leaf thermal tolerances are more closely adapted to microenvironment than macroclimate [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021]. Particularly, in an Australian arid region, *Acacia papyrocarpa* (Benth.) trees showed higher composite climate stress and greater $T_{50}$ in lower north-facing canopy than other canopy positions, correlating with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019]. However, more research is required to fully understand the relationship between leaf thermal tolerance and microenvironments.

*Further, if we take geographic gradients as a reliable proxy, we'd expect temperature sensitivities to reflect the environment to which the trees are adapted/ acclimatized.* Global $T_{50s}$ across latitude and climate range from ~45-50$^\circ$C (O' sullivan et al. 2017), where $T_{50}$ and $T_{crit}$ decrease with elevation among tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017].

*We're not aware of any other studies on this."we may want to argue that we have no *a priori* reasons to expect thermal tolerance to scale with height per se (or do we?). I think it makes sense to expect variation with light exposure because of the associated differences in temperature you can expect the leaves to be exposed to, but height itself should not affect heat tolerance other than through associated traits perhaps"--Martijn Slot*

Across species, $T_{50}$ is explained less by phylogeny [Slot et al. 2021; Perez and Feeley, 2020a] and more by leaf traits. $T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time [higher thermal time constant] whereas $T_{50}$ was greater among species' leaves with greater LMA (greater construction costs), compared to species with low LMA [@slotLeafHeatTolerance2020]. @perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that experienced higher maximum temperatures had higher photosynthetic heat tolerance. Therefore, we would expect, at vertical gradient, leaves exposed to greater heat stress to have higher heat tolerance but further research is required to confirm this.

@@

**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves (Chen et al. 2020), but temperature sensitivity of respiration--$Q_{10}$--shows no definite trend along the vertical gradient [@bolstad_foliar_1999; Weerasinghe et al.2014].**

>>>>

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen $1.2^\circ$C since the 1850-1900 period [@rohdeBerkeleyEarthLand2020] and are expected to rise over pre-industrial levels of $1.6 - 4.3^{\circ}$C by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilitiesa 2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilitiesa 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilitiesa 2021]. Collectively, these impacts may lead to reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020]. The resulting feedbacks to the climate system, decreased C storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016], yet this influence remains extremely uncertain [REFS].

@@

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
Foliage distribution is structured in correspondence with the distribution of tree sizes within a forest, where leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs. `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`). 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In open canopy forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations followed by disturbance all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
In this review, we focus on growing season conditions (peak LAD) unless otherwise noted.

@@

**Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al. 2015 ; Slot et al. 2019; Chen et al. 2020).**
This is primarily driven by the greater light available to sun-leaves, that also have traits (Table 1, Table 2) allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions. Photosynthesis increases with irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$. Light saturation levels differ between leaves and whole canopies with leaves saturating at lower light level than canopies; while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy.

@@

Leaf thermal tolerances are more closely adapted to microenvironment through leaf traits than macroclimate and phylogeny [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021]. Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ measured by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. At vertal gradient within canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower north-facing canopy than other canopy positions, correlating with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019]. Further, at a leaf trait level, $T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species' leaves with greater LMA, compared to species with low LMA [@slotLeafHeatTolerance2020]. Additionally, @perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that experienced higher maximum temperatures had higher photosynthetic heat tolerance. However, even with sun and shade leaf differences within canopy (Table 1)$T_{50}$ modestly varies, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019a]. Based on these evidences, we would expect, leaves exposed to greater heat stress in exposed canopies to have higher heat tolerance but further research is required to fully understand relationship between heat tolerance, light exposure and microclimates.

*If we take geographic gradients as a reliable proxy, we'd expect temperature sensitivities to reflect the environment to which the trees are adapted/ acclimatized.* Global $T_{50s}$ across latitude and climate range from ~45-50$^\circ$C (O' sullivan et al. 2017), where $T_{50}$ and $T_{crit}$ decrease with elevation among tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017].

@@

**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [@chenLeafEconomicsSpectrum2020], but temperature sensitivity of respiration--$Q_{10}$--shows no definite trend along the vertical gradient [@bolstad_foliar_1999; Weerasinghe et al.2014].**
